Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2021-01-18 Nicolas.tex
One column for referee!

\institute{Université de Lyon, F-69622, Lyon, France; Université de Lyon
    1, Villeurbanne; CNRS/IN2P3, Institut de Physique des Deux Infinis, Lyon
    \and 
    Université Clermont Auvergne, CNRS/IN2P3, Laboratoire de
    Physique de Clermont, F-63000 Clermont-Ferrand, France.
    \and
    Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720 
    \and
    Institut fur Physik, Humboldt-Universität zu Berlin, Newtonstr. 15,
    12489 Berlin
    \and
    University of Southampton: Southampton, GB
}

@@

\abstract{The detailed nature of type Ia supernovae (SNe~Ia) remains uncertain,
    and as survey statistics increase, the question of astrophysical
    systematic uncertainties rises, notably that of SN~Ia population
    evolution. In this paper, we study the dependence with redshift of the SN~Ia
    \texttt{SALT2.4} lightcurve stretch, a purely intrinsic SN property, to
    probe its potential redshift drift. The SN stretch has been shown to
    strongly correlate with the SN environment, notably with stellar age
    tracers. We model the underlying stretch distribution as a function of
    redshift, using the evolution of the fraction of young and old SNe~Ia as
    predicted by \cite{rigault2020}, and assuming constant underlying stretch
    distribution for each age population \textbf{consisting} of Gaussian
    mixtures. We test our prediction against published samples \textbf{cut} to
    have marginal magnitude selection effects, so that any observed change is
    indeed of astrophysical and not observational in origin. \textbf{In this first
    study, there are indications} that the underlying SN~Ia stretch distribution
    is evolving as a function of redshift, and that the young/old drifting model
    is a better description of the data than any time-constant model, including
    the sample-based asymmetric distributions often used to correct Malmquist
    bias. The favored underlying stretch model is the bimodal one derived from
    \cite{rigault2020}, \textbf{composed of} a high-stretch mode shared by both
    young and old environments, and a low-stretch mode exclusive to old
    environments. The precise impact of the redshift evolution of the SN Ia
    population intrinsic properties on cosmology remains to be studied. Yet, the
    astrophysical drift of the SN stretch distribution does affect current
    Malmquist bias corrections and thereby distances derived from SN affected by
observational selection effects. We highlight that such a bias will increase with surveys
covering increasingly larger redshift ranges, which is particularly important
for LSST.}

@@

Type Ia supernovae (SNe Ia) are powerful cosmological distance indicators that
enabled the discovery of the acceleration of the Universe's expansion
\citep{riess1998, perlmutter1999}. They remain today a key cosmological probe to
understand the properties of dark energy (DE) as it is the only tool able to
precisely map the recent expansion rate ($z<0.5$), when DE is driving it
\citep[e.g.][]{scolnicastro2020}. They also are key to directly measure the
Hubble Constant ($H_0$), provided one can calibrate their absolute magnitude
\citep{riess2016, freedman2019}. Interestingly, the value of $H_0$ derived when
the SNe~Ia are anchored to Cepheids \citep[the SH0ES project,][]{riess2009,
riess2016} is $\sim5\sigma$ higher than what is predicted from cosmic microwave
background (CMB) data measured by Planck assuming the standard $\L$CDM
\citep{planck2018, riess2019, reid2019}, or when the SN luminosity is anchored
at intermediate redshift by the baryon acoustic oscillation (BAO) scale
\citep{feeney2019}. While using the tip of the red giant branch technique in
place of the Cepheids seem to favor an intermediate value of $H_0$
\citep{freedman2019, freedman2020}, time delay measurements from strong lensing
seem to also favor high $H_0$ values \citep{wong2019}.

The $H_0$ tension has received a lot of attention, as it could be a sign of new
fundamental physics. Yet, no simple solution is able to accommodate this $H_0$
tension when accounting for all other probes \citep{knox2019}. Alternatively, systematic effects affecting one or several
of the aforementioned analyses might explain at least some of this tension.
\cite{rigault2015} suggested that SNe Ia from the Cepheid calibrator sample
differ by construction from those in the Hubble flow sample, as the former
strongly favors young stellar populations, while the latter does not. This
selection effect would impact the derivation of $H_0$ if SNe~Ia from young and
older environments differ in average standardized magnitudes.

%For the last decades, numerous analyses have studied 
The relation between SNe~Ia
and their host galaxy properties has been studied extensively. The first key finding
was that the standardized SNe~Ia magnitudes
significantly depend on the host galaxy stellar mass, with SNe~Ia from high-mass host 
galaxies being
brighter on average \cite[e.g.][]{kelly2010, sullivan2010, childress2013,
betoule2014, rigault2020, kim19}. This mass-step correction is currently used in
cosmological analyses \citep[e.g.][]{betoule2014, scolnic2018a}, including for
deriving $H_0$ \citep{riess2016, riess2019}. Yet, the underlying connection
between the SNe and their host galaxies remains unclear, particularly when using global properties such
as the host stellar mass. This in turn raises the question of the accuracy of such
corrections since host galaxy global properties evolve with redshift. 
More recently, studies have used the local SN host galaxy environment to probe
more direct connections between the SNe and their host galaxy environment \citep{rigault2013},
showing that local age tracers such as the Local specific Star Formation Rate
(LsSFR) or the local color are more strongly correlated with the standardized SN
magnitude \citep{roman2018, kim18,rigault2020}. Further reinforcing this
connection is the identification of SN~Ia spectral features that are 
correlated with LsSFR \citep{nordin2018}. These results suggest age as the driving
parameter underlying the mass-step. If true, this would have a significant
impact for cosmology, since the environmental correction to apply to SN
standardization could strongly vary with redshift. \citep{rigault2013,
childress2014, scolnic2018a}. Yet, the importance of local SN environmental
studies remains highly debated \cite[e.g.][]{jones2015, jones2019} and
especially the impact of such an astrophysical bias has on the derivation of
$H_0$ \citep{jones2015, riess2016, riess2018, rose2019}.

The concept of the SN~Ia age dichotomy arose with the study of the SN~Ia rate.
\cite{mannucci2005, scannapieco2005, sullivan2006, aubourg2008} have shown that
the relative SNe~Ia rate in galaxies could be explained if two populations
existed, one young, following the host star formation activity, and one old
following the host stellar mass (the so called ``prompt and delayed'' or ``A+B''
model). \cite{rigault2020} used the LsSFR to classify which are the
younger (those with a high LsSFR) and which are the older (those with low
LsSFR). Furthermore, since the first SNe~Ia host analyses, the SN stretch has
been known to be strongly correlated with the SN host galaxy properties
\citep{hamuy1996, hamuy2000}. This correlation that has been extensively confirmed
since \citep[e.g.][]{neill2009, sullivan2010, lampeitl2010, kelly2010,
gupta2011, dandrea2011, childress2013, rigault2013, pan2014, kim19}. Following
the ``A+B'' model and the connection between SN stretch and host galaxy properties,
\cite{howell2007} first discussed the potential redshift drift of the SN stretch
distribution. In this paper we revisit this question using the most recent SNe~Ia
datasets.

In this paper, we take a step aside \mr{from the cosmological analyses} to probe the validity of our modeling of the
SN population, which we claim to be constituted of two age-populations
\citep{rigault2013,rigault2015,rigault2020}: one old and one younger, the former
having on average lower lightcurve stretches and being brighter after
standardization. We use the correlation between the SN age, as probed by the
LsSFR, and the SN stretch to model the expected evolution of the underlying SN
stretch distribution as a function of redshift. This modeling relies on three
assumptions: (1) there are two distinct populations of SNe~Ia; (2) the relative
fraction of each of these populations as a function of redshift follows the
model presented in \cite{rigault2020} and (3) the underlying distribution of
stretch for each age sample is constant. This paper aims at testing this
specific model with datasets from the literature. 
% Note that the progenitor age
% as traced by the LsSFR seems to capture physical feature intrinsic to the
%progenitor and/or explosion mechanism that the stretch alone is not capturing
% \citep{nordin2018}.

@@

The ideal SN~Ia sample for studying this question would be a very deep, large-area,
volume-limited sample. This would capture the true underlying stretch distribution
function, and we would then study how it evolves with redshift. No such sample exists, 
so first we must construct subsamples that are as near to volume-limited as possible 
from existing high-redshift SN~Ia samples.

@@

We base our analysis on the most recent comprehensive SNe~Ia compilation, the
Pantheon catalog from \cite{scolnic2018a}. A naive approach to test the SN
stretch redshift drift would be to simply compare the observed SN stretch
distributions in a few bins of redshift. In practice, however, differential
observational selection effects will affect the observed SN stretch distributions. Indeed,
because the observed SN~Ia magnitude correlates with the lightcurve stretch (and
color), the first SNe~Ia that a magnitude-limited survey will miss are the
lowest-stretch (and reddest) ones. Consequently, if magnitude-related observational selection
effects are not accounted for, one might confuse true population drift with
survey properties, and conversely.

Assuming sufficient (and unbiased) spectroscopic follow-up for acquiring SN types
and host galaxy redshifts, the observation selection effects of magnitude-limited surveys should be
negligible below a given redshift at which even the faintest normal SNe~Ia can be
observed. \textbf{Aiding in the construction of nearly volume-limited 
subsamples is the fact that the SN~Ia population trails off towards fainter SNe~Ia.
A complication is that complete spectroscopic follow-up has not always been
the norm, as discussed below.}
In contrast, targeted surveys have highly complex observational
selection functions and so are discarded from our analysis. High-redshift SN cosmology
samples, such as Pantheon, are predominately from magnitude-limited surveys from
which volume-limited SN~Ia subsamples can be constructed.
%\textbf{which have easier to model observational selection functions}.

We present in Fig.~\ref{fig:maglim} the lightcurve stretch and color of SNe~Ia
from the following surveys: PanStarrs \citep[PS1][]{rest2014}, the Sloan Digital
Sky Survey \citep[SDSS][]{frieman2008} and the SuperNovae Legacy Survey
\citep[SNLS][]{astier2006}. An ellipse in the \textsc{\texttt{SALT2.4}} $(x_1,
c)$ plane with $x_1 = \pm 3$ and $c = \pm 0.3$ encapsulates the full
parent distribution \citep{guy2007,betoule2014}; see also \citet{bazin2011} and
\citet{campbell2013} for similar contours, the second using a more conservative
$|c| \leq 0.2$ cut. Assuming the SN absolute magnitude with $x_1=0$ and $c=0$ is
$M_0=-19.36$ \citep{kessler2009,scolnic2014}, we can derive the absolute
standardized magnitude at maximum of light $M = M_0 - \alpha x_1 + \beta c$
along the aforementioned ellipse given the standardization coefficient
$\alpha=0.156$ and $\beta=3.14$ from \cite{scolnic2018a}: the faintest SN~Ia is
that with $(x_1=-1.65, c=0.25)$ and an absolute standardized magnitude at peak
in Bessel $B$ band of $M^{t_0}_{\min} = -18.31$~mag. Since one ought to detect
this object typically 5 days before and a week after peak to build a suitable
lightcurve, the effective limiting standardized absolute magnitude is
approximately $M_{\lim} = -18.00$~mag. Hence, given the magnitude limit
$m_{\lim}$ of a magnitude limited survey, one can derive the maximum redshift
$z_{\lim}$ above which the faintest SNe~Ia will be missed using the relation
between apparent magnitude, redshift and absolute magnitude $\mu(z_{\lim}) =
m_{\lim} - M_{\lim}$.

@@

\textbf{We will thus consider a set of cuts that will define a first fiducial
    sample, taking the limits as initially suggested by the previous
    completeness analysis. \textbf{However,} as this solution might be an overly
    simplified way to create a complete sample (e.g., because it ignores spectroscopic follow-up
    in efficiency for redshifts below $z_{lim}$), we also consider another set of cuts to define a so-called
    ``conservative'' sample. The latter is smaller and therefore less statistically
    constraining, but also even less prone to observational selection effects. If the
    redshift drift is still significant in the conservative sample, it would be
    even more meaningful in a carefully-tailored selection-free sample. These
    samples are adequate for the goal
    of this study, which is to develop a first implementation of a model for drift in
    SN Ia properties. If fruitful, the sample selection can later be refined, if needed, with a
    more detailed
    %drift modeling before refining it on a more complex, yet more accurate,
%follow-up analysis also accounting for precise 
model of the observational selection, e.g., using
the SNANA package \citep{SNANA}.}

SNLS typically acquired SNe~Ia in the redshift range $0.4<z<0.8$; at these
redshifts, the rest-frame Bessel $B$ band roughly corresponds to the SNLS $i$
filter, which has a $5\sigma$ depth of
24.8~mag\footnote{\href{https://www.cfht.hawaii.edu/Science/CFHTLS/cfhtlsfinalreleaseexecsummary.html}{CFHT
final release website.}}. This converts to a $z_{\lim}=0.60$, in agreement with
\cite{neill2006}, \cite{perrett2010} and \cite{bazin2011}. Fig.~14 of
\citet[][see their Section~5]{perrett2010} suggests however a lower limit of
$z_{\lim}=0.55$. \textbf{We will therefore use $z=0.60$ and $z=0.55$ as redshift
limits for the fiducial and conservative samples, respectively, for SNLS.}

@@

In a similar redshift range, SDSS has a limiting magnitude of 22.5
\citep{dilday2008, sako2008}, which would lead to $z_{\lim}=0.24$. However, the
SDSS surveys \textbf{had to contend with limited spectroscopic resources. As
discussed in} \citet[][Section~2]{kessler2009}, during the first year of SDSS,
SNe~Ia with $r<20.5$~mag were favored for spectroscopic follow-up, corresponding
to a redshift cut at $0.15$. For the rest of the SDSS survey, additional
spectroscopic resources were available, and \cite{kessler2009} and
\cite{dilday2008} show a reasonable completeness up to $z_{\lim}=0.2$. Based on this,
\textbf{we will use $z_{\lim}=0.20$ and $z_{\lim}=0.15$ for our
fiducial and conservative samples, respectively, for SDSS.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/hist_surveys_cuts_55-cividis.pdf}
    \caption{\textit{From top to bottom}: Redshift histograms of SNe~Ia from the
        SDSS, PS1 and SNLS dataset respectively (data from Pantheon,
        \citealt{scolnic2018a}). The colored parts represent the distribution of
        SNe~Ia kept in our analysis for they are supposedly free from observational selection
    bias (see Section~\ref{sec:sample}). The darker (resp. lighter) color
responds to the conservative (resp. fiducial) selection cut.}
    \label{fig:cuts}
\end{figure}

@@

We report in Table~\ref{tab:bbc} \textbf{our determination of}
$\mu_0$ and $\sigma_{\pm}$ \textbf{for each sample when implementing an asymmetric Gaussian
model}, adjusted on the nominally selection-free samples using our fiducial cuts
(see Section~\ref{sec:sample}). We find our results in close agreement with
\cite{scolnic2016} for SNLS and SDSS and with \cite{scolnic2018a} for PS1, who
derived these model parameters using the full BBC formalism, using numerous
simulations to model the observational selection effects (see details e.g., Section~3 of
\citealt{kessler2017}). The agreement between our fit of the asymmetric
Gaussians on the supposedly selection-free part of the samples and the results
derived using the BBC formalism supports our approach to \textbf{constructing} a
sample with negligible observational selection effects. If we were to use \cite{scolnic2016}
and \cite{scolnic2018a} best fit values of the $\mu_0, \sigma_{\pm}$ asymmetric
parameters for SNLS, SDSS and PS1, respectively, the $\Delta$AIC between our
Base drifting model and the BBC modeling would go even deeper from $-32$ to
$-47$. We further discuss the consequence of this result for cosmology in
Section~\ref{sec:discussion}.

@@

Finally, ignoring the LsSFR measurements --- available only for the SNfactory
dataset (see Section~\ref{sec:modeling}) --- reduces the significance of the
results presented in this section, as expected. Even so, non-drifting models remain
strongly disfavored. For instance, the best fitting sample-based Gaussian
asymmetric model is still $\Delta\mathrm{AIC}<-10$ less representative of the
data than our Base drifting model.

@@

To the best of our knowledge, a SN~Ia stretch redshift drift modeling has never
been explicitly used in cosmological analyses, though Bayesian hierarchy
formalism such as UNITY \citep{rubin2015}, BAHAMAS \citep{shariff2016} or Steve
\citep{hinton2019} can easily allow it (see e.g., sections~1.3 and 2.5 of
\cite{rubin2015}). Not doing so is a second order issue for SN cosmology, as it
only affects the way one accounts for Malmquist bias. Indeed, as long as the
Phillips relation \citep{phillips1993} standardization parameter $\alpha$ is not
redshift dependent (a study \textbf{beyond} the scope of this paper, but see
e.g. \citealt{scolnic2018a}), the stretch-corrected SNe~Ia magnitudes used for
cosmology are blind to the underlying stretch distribution for complete samples.
However, surveys usually do have significant Malmquist bias for the upper half
of their SN redshift distribution. As a consequence, mismodeling of the
underlying stretch distribution will bias the SN magnitudes derived from such
surveys.

Commonly used Malmquist bias correction techniques, such as the BBC-formalism,
assume per sample asymmetric Gaussian functions to model the underlying
stretch and color distributions. Yet, as shown in Section~\ref{sec:results},
such a sample-based distribution is excluded in comparison to our drifting
model. Then, contrary to what \citet[][Section~2]{scolnic2016} and
\citet[][Section~5.4]{scolnic2018a} have suggested --- i.e., that traditional surveys
span sufficiently limited redshift ranges such that the per-sample approach accounts
for implicit redshift drifts --- a direct modeling of the redshift drift is more
appropriate than a sample-based approach. We \textbf{add} here that, as
measurements of modern surveys try to cover increasingly larger redshift ranges
in order to reduce calibration systematic uncertainties, this sample-based
approach \textbf{becomes} less valid, notably for PS1, DES and, soon, LSST.

>>>>

\institute{Univ Lyon, Univ Claude Bernard Lyon 1, CNRS, IP2I Lyon / IN2P3, IMR
    5822, F-69622, Villeurbanne, France
    \and 
    Université Clermont Auvergne, CNRS/IN2P3, Laboratoire de
    Physique de Clermont, F-63000 Clermont-Ferrand, France.
    \and
    Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720 
    \and
    Institut fur Physik, Humboldt-Universität zu Berlin, Newtonstr. 15,
    12489 Berlin
    \and
    University of Southampton: Southampton, GB
}

@@

\abstract{The \textbf{detailed} nature of type Ia supernovae (SNe~Ia) remains
    \textbf{uncertain}, and as survey statistics increase, the question of
    astrophysical systematic uncertainties rises, notably that of SN~Ia
    population evolution. In this paper, we study the dependence with redshift
    of the SN~Ia \texttt{SALT2.4} lightcurve stretch, a purely intrinsic SN
    property, to probe its potential redshift drift. The SN stretch has been
    shown to strongly correlate with the SN environment, notably with stellar
    age tracers. We model the underlying stretch distribution as a function of
    redshift, using the evolution of the fraction of young and old SNe~Ia as
    predicted by \cite{rigault2020}, and assuming constant underlying stretch
    distribution for each age population \textbf{consisting} of Gaussian
    mixtures. We test our prediction against published samples \textbf{cut} to
    have marginal magnitude selection effects, so that any observed change is
    indeed of astrophysical and not observational \textbf{in} origin. \textbf{In
    this first study, there are indications} that the underlying SN~Ia stretch
    distribution is evolving as a function of redshift, and that the young/old
    drifting model is a better description of the data than any time-constant
    model, including the sample-based asymmetric distributions \textbf{often}
    used to correct Malmquist bias. The favored underlying stretch model is the
    bimodal one derived from \cite{rigault2020}, \textbf{composed of} a
    high-stretch mode shared by both young and old environments, and a
    low-stretch mode exclusive to old environments. The precise impact of the
    redshift evolution of the SN Ia population intrinsic properties on cosmology
    remains to be studied. Yet, the astrophysical drift of the SN stretch
    distribution does affect current Malmquist bias corrections and thereby
distances derived from SN affected by \textbf{observational} selection effects.
We highlight that such a bias will increase with surveys covering increasingly
larger redshift ranges, which is particularly important for LSST.}

@@

Type Ia supernovae (SNe Ia) are powerful cosmological distance indicators that
enabled the discovery of the acceleration of the Universe's expansion
\citep{riess1998, perlmutter1999}. They remain today a key cosmological probe to
understand the properties of dark energy (DE) as it is the only tool able to
precisely map the recent expansion rate ($z<0.5$), when DE is driving it
\citep[e.g.][]{scolnicastro2020}. They also are key to directly measure the
Hubble Constant ($H_0$), provided one can calibrate their absolute magnitude
\citep{riess2016, freedman2019}. Interestingly, the value of $H_0$ derived when
the SNe~Ia are anchored \textbf{to} Cepheids \citep[the SH0ES
project,][]{riess2009, riess2016} is $\sim5\sigma$ higher than what is predicted
from cosmic microwave background (CMB) data measured by Planck assuming the
standard $\L$CDM \citep{planck2018, riess2019, reid2019}, or when the SN
luminosity is anchored at intermediate redshift by the baryon acoustic
oscillation (BAO) scale \citep{feeney2019}. While using the tip of the red giant
branch technique in place of the Cepheids seem to favor an intermediate value of
$H_0$ \citep{freedman2019, freedman2020}, time delay measurements from strong
lensing seem to also favor high $H_0$ values \citep{wong2019}.

The $H_0$ tension has received a lot of attention, as it could be a sign of new
fundamental physics. Yet, no simple solution is able to accommodate this $H_0$
tension when accounting for all other probes \citep{knox2019}. Alternatively,
systematic effects affecting one or several of the aforementioned analyses
\textbf{might} explain at least some of this tension. \cite{rigault2015}
suggested that SNe Ia from the Cepheid calibrator sample differ by construction
from \textbf{those in the Hubble flow sample,} as the former strongly favors
young stellar populations, while the latter \textbf{does} not. This selection
effect would impact the derivation of $H_0$ if SNe~Ia from young and older
environments differ in average standardized magnitudes.

\textbf{The relation between SNe~Ia and their host galaxy properties has been
studied extensively. The first key finding was} that the standardized SNe~Ia
magnitudes significantly depend on the host \textbf{galaxy} stellar mass, with
SNe~Ia from high-mass host \textbf{galaxies} being brighter on average
\cite[e.g.][]{kelly2010, sullivan2010, childress2013, betoule2014, rigault2020,
kim19}. This mass-step correction is currently used in cosmological analyses
\citep[e.g.][]{betoule2014, scolnic2018a}, including for deriving $H_0$
\citep{riess2016, riess2019}. Yet, the underlying connection between the SNe and
their host \textbf{galaxies} remains unclear\textbf{, particularly} when using
global properties such as the host stellar mass\textbf{. This in turn} raises
the question of the accuracy of \textbf{such corrections since host galaxy
global properties evolve with redshift}. More recently, studies have used the
local SN \textbf{host galaxy} environment to probe more direct connections
between the \textbf{SNe and their host galaxy environments} \citep{rigault2013},
showing that local age tracers such as the Local specific Star Formation Rate
(LsSFR) or the local color are more strongly correlated with the standardized SN
magnitude \citep{roman2018, kim18,rigault2020}. \textbf{Further reinforcing this
connection is the identification of SN~Ia spectral features that are correlated
with LsSFR \citep{nordin2018}. These results suggest} age as the driving
parameter underlying the mass-step. If true, this would have a significant
impact for cosmology, since the environmental correction to apply to SN
standardization could strongly vary with redshift. \citep{rigault2013,
childress2014, scolnic2018a}. Yet, the importance of local SN environmental
studies remains highly debated \cite[e.g.][]{jones2015, jones2019} and
especially the impact of such an astrophysical bias has on the derivation of
$H_0$ \citep{jones2015, riess2016, riess2018, rose2019}.

The concept of the SN~Ia age dichotomy arose with the study of the SN~Ia rate.
\cite{mannucci2005, scannapieco2005, sullivan2006, aubourg2008} have shown that
the relative SNe~Ia rate in galaxies could be explained if two populations
existed, one young, following the host star formation activity, and one old
following the host stellar mass (the so called ``prompt and delayed'' or ``A+B''
model). \cite{rigault2020} used the LsSFR to classify which are the younger
(those with a high LsSFR) and which are the older (those with low LsSFR).
Furthermore, since the first SNe~Ia host analyses, the SN stretch has been known
to be strongly correlated with the SN host \textbf{galaxy} properties
\citep{hamuy1996, hamuy2000}. This correlation that has been extensively
confirmed since \citep[e.g.][]{neill2009, sullivan2010, lampeitl2010, kelly2010,
gupta2011, dandrea2011, childress2013, rigault2013, pan2014, kim19}. Following
the ``A+B'' model and the connection between SN stretch and host galaxy
properties, \cite{howell2007} first discussed the potential redshift drift of
the SN stretch distribution. In this paper we revisit this \textbf{question
using} the most recent SNe~Ia \textbf{datasets}.

In this paper, we take a step aside \textbf{from the cosmological analyses} to
probe the validity of our modeling of the SN population, which we claim to be
constituted of two age-populations \citep{rigault2013,rigault2015,rigault2020}:
one old and one younger, the former having on average lower lightcurve stretches
and being brighter after standardization. We use the correlation between the SN
age, as probed by the LsSFR, and the SN stretch to model the expected evolution
of the underlying SN stretch distribution as a function of redshift. This
modeling relies on three assumptions: (1) there are two distinct populations of
SNe~Ia; (2) the relative fraction of each of these populations as a function of
redshift follows the model presented in \cite{rigault2020} and (3) the
underlying distribution of stretch for each age sample is constant. This paper
aims at testing this specific model with datasets from the literature.

@@

\textbf{The ideal SN~Ia sample for studying this question would be a very deep,
    large-area, volume-limited sample. This would capture the true underlying
    stretch distribution function, and we would then study how it evolves with
    redshift. No such sample exists, so first we must construct subsamples that
are as near to volume-limited as possible from existing high-redshift SN~Ia
samples.}

@@

We base our analysis on the most recent comprehensive SNe~Ia compilation, the
Pantheon catalog from \cite{scolnic2018a}. A naive approach to test the SN
stretch redshift drift would be to simply compare the observed SN stretch
distributions in a few bins of redshift. In practice, however, differential
\textbf{observational} selection effects \textbf{will affect} the observed SN
stretch distributions. Indeed, because the observed SN~Ia magnitude correlates
with the lightcurve stretch (and color), the first SNe~Ia that a
magnitude-limited survey will miss are the lowest-stretch (and reddest) ones.
Consequently, if magnitude-related \textbf{observational} selection effects are
not accounted for, one might confuse true population drift with survey
properties, and conversely.

Assuming sufficient (and unbiased) spectroscopic follow-up for acquiring
\textbf{SN types and host galaxy redshifts}, the \textbf{observation} selection
effects of magnitude-limited surveys should be negligible below a given redshift
at which even the faintest \textbf{normal} SNe~Ia can be observed.
\textbf{Aiding in the construction of nearly volume-limited subsamples is the
    fact that the SN~Ia population trails off towards fainter SNe~Ia. A
    complication is that complete spectroscopic follow-up has not always been
    the norm, as discussed below.} In contrast, targeted surveys have highly
    complex \textbf{observational} selection functions and \textbf{so are}
    discarded from our analysis. \textbf{High-redshift SN cosmology samples,
    such as Pantheon, are predominately from magnitude-limited surveys from
which volume-limited SN~Ia subsamples can be constructed.}

We present in Fig.~\ref{fig:maglim} the lightcurve stretch and color of SNe~Ia
from the following surveys: PanStarrs \citep[PS1][]{rest2014}, the Sloan Digital
Sky Survey \citep[SDSS][]{frieman2008} and the SuperNovae Legacy Survey
\citep[SNLS][]{astier2006}. An ellipse in the \textsc{\texttt{SALT2.4}} $(x_1,
c)$ plane with $x_1 = \pm 3$ and $c = \pm 0.3$ encapsulates the full
\textbf{parent} distribution \citep{guy2007, betoule2014}; see also
\citet{bazin2011} and \citet{campbell2013} for similar contours, the second
using a more conservative $|c| \leq 0.2$ cut. Assuming the SN absolute magnitude
with $x_1=0$ and $c=0$ is $M_0=-19.36$ \citep{kessler2009,scolnic2014}, we can
derive the absolute standardized magnitude at maximum of light $M = M_0 - \alpha
x_1 + \beta c$ along the aforementioned ellipse given the standardization
coefficient $\alpha=0.156$ and $\beta=3.14$ from \cite{scolnic2018a}: the
faintest SN~Ia is that with $(x_1=-1.65, c=0.25)$ and an absolute standardized
magnitude at peak in Bessel $B$ band of $M^{t_0}_{\min} = -18.31$~mag. Since one
ought to detect this object typically 5 days before and a week after peak to
build a suitable lightcurve, the effective limiting standardized absolute
magnitude is approximately $M_{\lim} = -18.00$~mag. Hence, given the magnitude
limit $m_{\lim}$ of a magnitude limited survey, one can derive the maximum
redshift $z_{\lim}$ above which the faintest SNe~Ia will be missed using the
relation between apparent magnitude, redshift and absolute magnitude
$\mu(z_{\lim}) = m_{\lim} - M_{\lim}$.

@@

\textbf{We will thus consider a set of cuts that will define a first fiducial
    sample, taking the limits as initially suggested by the previous
    completeness analysis. However, as this solution might be an overly
    simplified way to create a complete sample, e.g., because it ignores
    spectroscopic follow-up in efficiency for redshifts below $z_{lim}$, we
    also consider another set of cuts to define a so-called ``conservative''
    sample. The latter is smaller and therefore less statistically constraining,
    but also even less prone to observational selection effects. If the redshift
    drift is still significant in the conservative sample, it would be even more
    meaningful in a carefully-tailored selection-free sample. These samples are
    adequate for the goal of this study, which is to develop a first
    implementation of a model for drift in SN Ia properties. If fruitful, the
    sample selection can later be refined, if needed, with a more detailed model
of the observational selection, e.g., using the SNANA package \citep{SNANA}.}

SNLS typically acquired SNe~Ia in the redshift range $0.4<z<0.8$; at these
redshifts, the rest-frame Bessel $B$ band roughly corresponds to the SNLS $i$
filter, \textbf{which} has a $5\sigma$ depth of
24.8~mag\footnote{\href{https://www.cfht.hawaii.edu/Science/CFHTLS/cfhtlsfinalreleaseexecsummary.html}{CFHT
final release website.}}. This converts to a $z_{\lim}=0.60$, in agreement with
\cite{neill2006}, \cite{perrett2010} and \cite{bazin2011}. Fig.~14 of
\citet[][see their Section~5]{perrett2010} suggests however a lower limit of
$z_{\lim}=0.55$. \textbf{We will therefore use $z=0.60$ and $z=0.55$ as redshift
limits for the fiducial and conservative samples, respectively, for SNLS.}

@@

In a similar redshift range, SDSS has a limiting magnitude of 22.5
\citep{dilday2008, sako2008}, which would lead to $z_{\lim}=0.24$. However, the
SDSS surveys \textbf{had to contend with limited spectroscopic resources. As
discussed in} \citet[][Section~2]{kessler2009}, during the first year of SDSS,
SNe~Ia with $r<20.5$~mag were favored for spectroscopic follow-up, corresponding
to a redshift cut at $0.15$. For the rest of the SDSS survey, additional
spectroscopic resources were available, and \cite{kessler2009} and
\cite{dilday2008} show a reasonable completeness up to $z_{\lim}=0.2$.
\textbf{Based on this, we will use $z_{\lim}=0.20$ and $z_{\lim}=0.15$ for our
fiducial and conservative samples, respectively, for SDSS.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/hist_surveys_cuts_55-cividis.pdf}
    \caption{\textit{From top to bottom}: Redshift histograms of SNe~Ia from the
        SDSS, PS1 and SNLS dataset respectively (data from Pantheon,
        \citealt{scolnic2018a}). The colored parts represent the distribution of
        SNe~Ia kept in our analysis for they are supposedly free from
    \textbf{observational} selection bias (see Section~\ref{sec:sample}). The
darker (resp. lighter) color responds to the conservative (resp. fiducial)
selection cut.}
    \label{fig:cuts}
\end{figure}

@@

We report in Table~\ref{tab:bbc} \textbf{our determination of} $\mu_0$ and
$\sigma_{\pm}$ \textbf{for each sample when implementing an asymmetric Gaussian
model}, adjusted on the nominally selection-free samples using our fiducial cuts
(see Section~\ref{sec:sample}). We find our results in close agreement with
\cite{scolnic2016} for SNLS and SDSS and with \cite{scolnic2018a} for PS1, who
derived these model parameters using the full BBC formalism, using numerous
simulations to model the \textbf{observational} selection effects (see details
e.g., Section~3 of \citealt{kessler2017}). The agreement between our fit of the
asymmetric Gaussians on the supposedly selection-free part of the samples and
the results derived using the BBC formalism supports our approach to
\textbf{constructing} a sample with negligible \textbf{observational} selection
effects. If we were to use \cite{scolnic2016} and \cite{scolnic2018a} best fit
values of the $\mu_0, \sigma_{\pm}$ asymmetric parameters for SNLS, SDSS and
PS1, respectively, the $\Delta$AIC between our Base drifting model and the BBC
modeling would go even deeper from $-32$ to $-47$. We further discuss the
consequence of this result for cosmology in Section~\ref{sec:discussion}.

@@

Finally, ignoring the LsSFR measurements --- available only for the SNfactory
dataset (see Section~\ref{sec:modeling}) --- reduces the significance of the
results presented in this section, as expected. \textbf{Even so}, non-drifting
models remain strongly disfavored. \textbf{For} instance, the best
\textbf{fitting} sample-based Gaussian asymmetric \textbf{model} is still
$\Delta\mathrm{AIC}<-10$ less representative of the data than our Base drifting
\textbf{model}.

@@

To the best of our knowledge, a SN~Ia stretch redshift drift modeling has never
been explicitly used in cosmological analyses, though Bayesian hierarchy
formalism such as UNITY \citep{rubin2015}, BAHAMAS \citep{shariff2016} or Steve
\citep{hinton2019} can easily allow it (see e.g., sections~1.3 and 2.5 of
\cite{rubin2015}). Not doing so is a second order issue for SN cosmology, as it
only affects the way one accounts for Malmquist bias. Indeed, as long as the
Phillips relation \citep{phillips1993} standardization parameter $\alpha$ is not
redshift dependent (a study \textbf{beyond} the scope of this paper, but see
e.g. \citealt{scolnic2018a}), the stretch-corrected SNe~Ia magnitudes used for
cosmology are blind to the underlying stretch distribution for complete samples.
However, surveys usually do have significant Malmquist bias for the upper half
of their SN redshift distribution. As a consequence, \textbf{mismodeling} of the
underlying stretch distribution will bias the SN magnitudes derived from such
surveys.

Commonly used Malmquist bias correction techniques, such as the BBC-formalism,
assume per sample asymmetric Gaussian functions \textbf{to} model the underlying
stretch and color distributions. Yet, as shown in Section~\ref{sec:results},
such a sample-based distribution is excluded \textbf{in comparison to} our
drifting model. Then, \textbf{contrary to} what \citet[][Section~2]{scolnic2016}
and \citet[][Section~5.4]{scolnic2018a} have suggested --- i.e., that
traditional surveys span \textbf{sufficiently} limited redshift ranges
\textbf{such that} the per-sample approach accounts for implicit redshift drifts
--- a direct modeling of the redshift drift is more appropriate than a
sample-based approach. We \textbf{add} here that, as measurements of modern
surveys try to cover increasingly larger redshift ranges in order to reduce
calibration systematic uncertainties, this sample-based approach
\textbf{becomes} less valid, notably for PS1, DES and, soon, LSST.
